---
title: "PA1_template.Rmd"
author: "David"
date: "16 December 2015"
output: html_document
---

This is my analysis of the activity data for the Reproducible Research course, assignment 1.


The first task is to load the data from the CSV file. i set the working directory to a local drive.

```{r}

setwd("C:/David/data/repdata-035")

myDataSet <- read.csv("activity.csv")

myDataSet$date <- as.character(myDataSet$date)

head(myDataSet)


```

First step is to show a histogram of this data to show the total number of steps taken. I will omit the NAs here, and aggregate the data on the data to be able to show the total number of steps taken in the day.

```{r}

myDailyData <- na.omit(myDataSet)

head(myDailyData)

myDailyData <- aggregate(myDataSet$steps, by=list(myDataSet$date), FUN=sum)

hist(myDailyData[,2], breaks=20, main="Total number of steps per day", col="blue", xlab="Steps taken")


```



Using the same aggregated data I will show the mean...

```{r}

mean(myDailyData[,2], na.rm=TRUE)

```

...and the median.

```{r}

median(myDailyData[,2], na.rm=TRUE)

```


This is the daily pattern of the steps taken during a day. The interval measurements are means.

```{r}

ActiveDailyPattern <- aggregate(steps ~ interval, na.omit(myDataSet), mean)

head(ActiveDailyPattern)

plot(
        ActiveDailyPattern$interval,
        ActiveDailyPattern$steps,
        type = "l",
        main = "Active Daily Pattern",
        xlab = "5-Minute Interval",
        ylab = "Average Number of Steps Taken"
)


```

The 5-minute interval which, on average across all the days in the dataset, which contains the maximum number of steps is...

```{r}
BiggestInterval <- ActiveDailyPattern[ActiveDailyPattern$steps==max(ActiveDailyPattern$steps), ]

BiggestInterval[1,1]

```


*Imputing missing values*

First of all count the number of NAs in the dataset.


```{r}

nas<-sum(is.na(myDataSet))

nas

```

Now we are going to fill in the NA items with the equivilant average item for the interval across the whole dataset.

```{r}

CompleteDataSet <- myDataSet
CompleteDataSet$steps[which(is.na(CompleteDataSet$steps))] <- tapply(CompleteDataSet$steps, CompleteDataSet$interval, mean, na.rm=TRUE, simplify=FALSE )
nas<-sum(is.na(CompleteDataSet))
nas

```

Using this data i will make a histogram of the total number of steps taken each day

```{r}

CompleteDataSet$steps <- as.vector(CompleteDataSet$steps, mode="numeric")

CompleteDataSetAggregated <- aggregate(CompleteDataSet$steps, by=list(CompleteDataSet$date), FUN=sum)

hist(CompleteDataSetAggregated[,2], breaks=20, main="Total number of steps per day", col="green", xlab="Steps taken")


```

Calculate and report the mean  total number of steps taken per day. 

```{r}

mean(CompleteDataSetAggregated$x, na.rm = TRUE)


```

Calculate and report the median total number of steps taken per day. 

```{r}


median(CompleteDataSetAggregated$x, na.rm = TRUE)


```

As can be seen the mean and the median are now the same, This is the impact of imputting the NA values.

*Differences in activity patterns between weekdays and weekends*

New factor variable created in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}

CompleteDataSetDays <- CompleteDataSet
CompleteDataSetDays$day <- weekdays(as.Date(CompleteDataSet$date))
WeekDayConvert <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
        "weekend"
    } else {
        "weekday"
    }
}
CompleteDataSetDays$day <- as.factor(sapply(CompleteDataSetDays$date, WeekDayConvert))
head(CompleteDataSetDays)
```

A panel plot created containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}

Weekends <- subset(CompleteDataSetDays, CompleteDataSetDays$day == "weekend")
Weekday <- subset(CompleteDataSetDays, CompleteDataSetDays$day == "weekday")

Weekends <- aggregate(steps ~ interval, na.omit(Weekends), mean)
Weekday <- aggregate(steps ~ interval, na.omit(Weekday), mean)


par(mfrow=c(2,1))

plot(
        Weekends$interval,
        Weekends$steps,
        type = "l",
        main = "Active Weekend Pattern",
        xlab = "5-Minute Interval",
        ylab = "Ave. No. of Steps"
)


plot(
        Weekday$interval,
        Weekday$steps,
        type = "l",
        main = "Active Weekday Pattern",
        xlab = "5-Minute Interval",
        ylab = "Ave. No. of Steps"
)


```



